A Simple Ratio for Estimating the Effective Reproduction Value of an Epidemic

### 2021-01-16 docs/manuscript/manuscript.Rmd
manuscript writing

# Convert number to text 
# Ref: https://5harad.com/mse125/r/visualization_code.html
addUnits <- function(n) {
    labels <- ifelse(n < 1000, n,  # less than thousands
                     ifelse(n < 1e6, paste0(round(n/1e3), ' thousand'),  # in thousands
                            ifelse(n < 1e9, paste0(round(n/1e6), ' million'),  # in millions
                                   ifelse(n < 1e12, paste0(round(n/1e9), ' billion'), # in billions
                                          'too big!'
                                   ))))
    return(labels)
}

@@

  englishList = paste(head(x, -1), collapse = ", ")
  englishList = paste(englishList, last_item, sep = ", and ")

@@

N = simulation_parameters$initial_state_values %>% sum()
error_median = with(rt_estimates, median(abs(rt.simple_mean_error), na.rm = T)) %>% signif(digits = 2)
error_quantile_975 = with(rt_estimates, quantile(abs(rt.simple_mean_error), 0.975, na.rm = T)) %>% signif(digits = 2) %>% format(scientific = F)
error_quantile_025 = with(rt_estimates, quantile(abs(rt.simple_mean_error), 0.025, na.rm = T)) %>% signif(digits = 2) %>% format(scientific = F)

national_data_filtered = national_data %>% 
  filter(time > as.Date("2020-04-01") & time < as.Date("2020-12-01")) %>% 
  filter(cases_new > 19)

@@

# Introduction {-}

The effective reproduction number ($R_t$) is a fundamental measure in infectious disease epidemiology. It can be defined as the average number of secondary caused by an infected individual. The $R_t$ changes over the course of an epidemic, because it is not only determined by characteristics of the pathogen but also by human behavior (e.g. face mask usage, physical distancing, and travel restrictions). If the $R_t$ is above 1, it indicates that epidemic is growing exponentially. If, by contrast, the $R_t$ drop below 1, it indicates that the number of new cases is decreasing. Furthermore, the $R_t$ can vary between geographic locations, as social behaviors vary between these locations.

From a public health perspective, the $R_t$ can be a method to communicate whether an epidemic is growing in a region. However, this communication may be hindered by the fact that many of the current methods for estimating $R_t$ require advanced mathematics to understand [@cori_new_2013; @wallinga_different_2004; @wallinga_how_2007]. Here, we describe how $R_t$ can be approximated by a ratio of new infections on day $t$ and day $t-d$, where d is the mean generation interval of the infection ("generation interval" is defined as the average time from when a primary case becomes infectious to when a secondary becomes infectious).

$$R_t \approx \frac{I_t}{I_{t-d}}$$

The major advantage of this approximation is that its simplicity makes it easily understandable and verifiable. Anybody with basic math skills and access to the new case counts can calculate this ratio to estimate the $R_t$.

## A Simple Ratio for Estimating Rt {-}

The total number of new infections at time, $I_t$, can be calculated by summing the expected number of new cases created at $t$ by each person infected prior to $t$. The expected of number of new cases created by an infected person can be considered as a function, $n(a)$, of the age ($a$) of their infection [@wallinga_how_2007].

\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}n(a)}da
  (\#eq:eq2)
\end{equation}

The function $n(a)$ can be normalized to by the total number of infection an infected person is expected cause (i.e. $R$) to produce the generation interval distribution:

\begin{equation} 
  g(a) = \frac{n(a)}{R_t}
  (\#eq:eq2)
\end{equation}

Equation (2) can be substituted into Equation (1) to produce:

\begin{equation} 
  \frac{I_t}{R_t} = \int_a^\infty{I_{t-a}g(a)}da
  (\#eq:eq3)
\end{equation}

\begin{equation} 
  R_t = \frac{I_t}{\int{I_{t-a}g(a)}da}
  (\#eq:eq4)
\end{equation}

If we make the assumption that there is minimal variance in the generation interval distribution (i.e. $g(a)$ equals approximately 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ equals approximately zero for all other values of $a$), then equation (4) can be simplified to:

\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}

We can make this estimation of $R_t$ robust to variations in the generation interval and reporting of the infection by calculating $R_t$ as the ratio of the sums of new cases within two time windows size $\tau$ days (with the center of the time periods being on days $t$ and $t-d$):

\begin{equation} 
  R_t \approx \frac{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_{i-d}} = \frac{n_{t}}{n_{t-d}}
  (\#eq:eq6)
\end{equation}

A confidence interval can be generated for this ratio by carrying forward a confidence interval in the generation interval.

# Materials and methods {-}

## Simulation {-}

@@

## Rt Esimation {-}

@@

## COVID-19 Data {-}

@@

## Code and Data Availability {-}

All simulations, data processing, analysis, and graphics were all created using the R statistical software using the following packages: `r vector_to_englishList(packages)` [`r package_citation_ids`]. All code and data used to develop this project is available at [https://github.com/aryazand/simple_rt](https://github.com/aryazand/simple_rt)

# Results {-}

## Accuracy of the simple ratio estimation of $R_t$ {-}

@@

To assess accuracy of $R_t$ estimations, we used Susceptible-Exposed-Infectious-Resolved (SEIR) model to simulate epidemics and then compare how accurately the two methods can estimate the $R_t$ values in the epidemics. We simulated `r simulation_parameters$Nsims` epidemics in a population of `r addUnits(N)` individuals. We varied the lengths of latent and infectiousness periods with each simulation. In each simulation, the population underwent changes in the $R_t$ value representing changes in population behavior (e.g. "lockdown" period or lifting of a lockdown).  **Figure \@ref(fig:sim-results) A** shows the true $R_t$ and the estimated $R_t$ values for each day in one such simulated epidemic. A full visualization of the all the simulations can be seen in (**Figure \@ref(fig:all-simulations)**).

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was  `r error_median` (90% range of `r error_quantile_025` to `r error_quantile_975`) (**Figure \@ref(fig:sim-results) B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (**Figure \@ref(fig:sim-results) C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007].

@@

panel_A <- rt_estimates %>%
  filter(simulation_run == 66) %>%
  create_estimate_plot()

panel_B <- rt_estimates %>%
  filter(time > 14) %>%
  create_rt_estimation_error_boxplot()

panel_C <- rt_estimates %>%
  filter(time > 14) %>%
  create_rt_estimation_error_by_rt()

ggarrange(panel_A, panel_B, panel_C, ncol=3, nrow=2, align = "hv", 
          labels=c("A", "B", "C", "D", "E", "F"), common.legend = T)
```

Often however, the the mean and standard deviation of the generation interval is typically not known exactly. For instance, for the COVID-19 the mean generation interval estimates, with 95% credible intervals, include 3.95 (3.01 - 4.91) and 5.20 (3.78 - 6.78); and estimates of the standard deviation of the generation interval include 1.51 (0.74 - 2.97) and 1.72 (0.91 - 3.93) [@ganyani_estimating_2020]. Therefore, we compared the accuracy of the two methods if the standard deviation of the generation interval is mis-specified as three days longer. Under these conditions, the simple ratio actually performs slightly better than than the Cori method (**Figure \@ref(fig:estimation-inaccParams)**).

@@

## Using the Simple Ratio to Estimate $R_t$ with real-world COVID-19 data {-}

We next compared $R_t$ estimates from the Simple Ratio and Cori et al for using data from the current COVID-19 pandemic. We estimated $R_t$ in four nations using a generation interval with mean of `r COVID19_params$mean_gi` and standard deviation of `r COVID19_params$sd_gi`[@ganyani_estimating_2020]. **Figure \@ref(fig:real-world-data) A** shows the results for the month of October 2020. Visual inspection shows that the two methods of estimating $R_t$ have very good concordance. Across all nations over the course of the pandemic, the median difference in estimation between the two methods was `r estimation_diff_median` (95% range of `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`) (**Figure \@ref(fig:real-world-data) B**).

# Discussion {-}

# References {-}

<div id="refs"></div>

# Supplementary Material {-}
\beginsupplement

\newpage
\blandscape
```{r all-simulations, fig.width = 11, fig.height = 6, fig.cap = "All Simulations"}

rt_estimates %>%
  create_estimate_plot() + 
  facet_wrap(~simulation_run) + 
  theme(strip.background = element_blank(), strip.text = element_blank())
```
\elandscape

\newpage
```{r estimation-inaccParams,  fig.cap = "(A) estimation with gi mean mis-specified. (B) estimation with gi sd mis-specified."}

>>>>

# Convert number to text 
# Ref: https://5harad.com/mse125/r/visualization_code.html
addUnits <- function(n) {
  labels <- ifelse(n < 1000, n,  # less than thousands
                   ifelse(
                     n < 1e6,
                     paste0(round(n / 1e3), ' thousand'),
                     # in thousands
                     ifelse(
                       n < 1e9,
                       paste0(round(n / 1e6), ' million'),
                       # in millions
                       ifelse(n < 1e12, paste0(round(n / 1e9), ' billion'), # in billions
                              'too big!')
                     )
                   ))
  return(labels)
}

@@

  englishList = paste(head(x,-1), collapse = ", ")
  englishList = paste(englishList, last_item, sep = ", and ")

@@

# Comparison with COVID-19 data
national_data_filtered = national_data %>% 
  filter(time > as.Date("2020-04-01") & time < as.Date("2020-12-01")) %>% 
  filter(cases_new > 19)

@@

# Introduction {.unnumbered}

The effective reproduction number ($R_t$) is a fundamental measure in infectious disease epidemiology. It can be defined as the average number of secondary cases caused by an infected individual. The $R_t$ changes over the course of an epidemic, because it is not only determined by characteristics of the pathogen but also by human behavior (e.g. face mask usage, physical distancing, and travel restrictions). If the $R_t$ is above 1, it indicates that the epidemic is growing exponentially. If, by contrast, the $R_t$ drop below 1, it indicates that the number of new cases is decreasing. Furthermore, the $R_t$ can vary between geographic locations, as social behaviors and environmental factors vary between these locations.

From a public health perspective, the $R_t$ can be a method to communicate whether an epidemic is growing in a region. However, this communication may be hindered by the fact that many of the current methods for estimating $R_t$ require advanced mathematics to understand and to apply [@cori_new_2013; @wallinga_different_2004; @wallinga_how_2007]. Here, we describe how $R_t$ can be approximated by a ratio of new infections on day $t$ and day $t-d$, where $d$ is the mean generation interval of the infection (Equation 1). The average *generation interval* is defined as the average time from when a primary case becomes infectious to when a secondary case becomes infectious.

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
The major advantage of this approximation is that its simplicity makes it easily understandable and verifiable. Anybody with basic math skills and access to the new case counts can calculate this ratio to estimate the $R_t$.

## A Simple Ratio for Estimating Rt {.unnumbered}

As described by previous publications [@wallinga_how_2007], we can define a function $n(a)$ that defines the expected number of new infections created today (time $t$) by a single person who was infected $a$ days ago (at time $t-a$). Therefore, to calculate $N_a$ (the total expected number of new infections created today by all individuals who became infected $a$ days ago), we multiply $n(a)$ by the number of new infections created $a$ days ago ($I_{t-a}$).

```{=tex}
\begin{equation}
  N_a = I_{t-a}n(a)
  (\#eq:eq2)
\end{equation}
```
Then to determine the total expected number of new infections today ($I_t$), we sum $N_a$ for all values of $a$ (i.e. sum up the expected number of new infections created by each infected individuals based on the age of their infection). If we treat $a$ and $t$ instead of discrete days, we integrate over all values of $a$ to produce Equation \@ref(eq:renewal-eq).

```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}n(a)}da
  (\#eq:renewal-eq)
\end{equation}
```
The generation interval distribution is equal to $n(a)$ normalized by the total number of infections a single infected person is expected cause (i.e. $R_t$).

```{=tex}
\begin{equation} 
  g(a) = \frac{n(a)}{R_t}
  (\#eq:generation-interval-eq)
\end{equation}
```
Thus, Equation \@ref(eq:generation-interval-eq) can be substituted into Equation \@ref(eq:renewal-eq) to produce:

```{=tex}
\begin{equation} 
  \frac{I_t} = \int_a^\infty{I_{t-a}\frac{g(a)}{R_t}}da
  (\#eq:sub-in-rt-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  \frac{I_t}{R_t} = \int_a^\infty{I_{t-a}g(a)}da
  (\#eq:sub-in-rt-2)
\end{equation}
```
```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\int{I_{t-a}g(a)}da}
  (\#eq:sub-in-rt-3)
\end{equation}
```
If we make the assumption that there is minimal variance in the generation interval distribution (i.e. $g(a)$ equals approximately 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ equals approximately zero for all other values of $a$), then equation (4) can be simplified to:

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
We can make this estimation of $R_t$ robust to variations in the generation interval and reporting of new cases by calculating $R_t$ as the ratio of the sums of new cases within two time windows size $\tau$ days (with the center of the time windows being on days $t$ and $t-d$):

```{=tex}
\begin{equation} 
  R_t \approx \frac{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_{i-d}} = \frac{n_{t}}{n_{t-d}}
  (\#eq:eq6)
\end{equation}
```
A confidence interval can be generated for this ratio by carrying forward a confidence interval in the generation interval.

# Materials and methods {.unnumbered}

## Simulation {.unnumbered}

@@

## Rt Estimation {.unnumbered}

@@

## COVID-19 Data {.unnumbered}

@@

## Code and Data Availability {.unnumbered}

All simulations, data processing, analysis, and graphics were all created using the R statistical software using the following packages: `r vector_to_englishList(packages)` [`r package_citation_ids`]. All code and data used to develop this project is available at <https://github.com/aryazand/simple_rt>

# Results {.unnumbered}

## Accuracy of the simple ratio estimation of $R_t$ {.unnumbered}

@@

To assess accuracy of $R_t$ estimations, we used Susceptible-Exposed-Infectious-Resolved (SEIR) model to simulate epidemics and then compare how accurately the two methods can estimate the $R_t$ values in the epidemics. We simulated `r simulation_parameters$Nsims` epidemics in a population of `r addUnits(N)` individuals. We varied the lengths of latent and infectiousness periods with each simulation. In each simulation, the population underwent changes in the $R_t$ value representing changes in population behavior (e.g. "lockdown" period or lifting of a lockdown). **Figure** \@ref(fig:sim-results) **A** shows the true $R_t$ and the estimated $R_t$ values for each day in one such simulated epidemic. A full visualization of the all the simulations can be seen in (**Figure** \@ref(fig:all-simulations)).

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_median` (90% range of `r error_quantile_025` to `r error_quantile_975`) (**Figure** \@ref(fig:sim-results) **B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (**Figure** \@ref(fig:sim-results) **C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007].

@@

# Descriptive Statistics
N = simulation_parameters$initial_state_values %>% sum()
error_median = with(rt_estimates, median(abs(rt.simple_mean_error), na.rm = T)) %>% signif(digits = 2)
error_quantile_975 = with(rt_estimates, quantile(abs(rt.simple_mean_error), 0.975, na.rm = T)) %>% signif(digits = 2) %>% format(scientific = F)
error_quantile_025 = with(rt_estimates, quantile(abs(rt.simple_mean_error), 0.025, na.rm = T)) %>% signif(digits = 2) %>% format(scientific = F)

# 
rt_estimates_error_summary <- rt_estimates %>%
  filter(time > 14) %>%
  group_by(simulation_run) %>%
  summarise(dm_test = list(dm.test(rt.simple_mean_error, rt.cori_mean_error)))

# Plots
panel_A <- rt_estimates %>%
  filter(simulation_run == 66) %>%
  create_estimate_plot()

panel_C <- rt_estimates %>%
  filter(time > 14) %>%
  create_rt_estimation_error_boxplot()

panel_D <- rt_estimates %>%
  filter(time > 14) %>%
  create_rt_estimation_error_by_rt()

panel_D <- rt_estimates %>% 
  pivot_longer(cols = ends_with("mean_error"), names_to = "Method", values_to = "estimation_error") %>% 
  mutate(Method = recode(Method, rt.cori_mean_error = "cori", rt.simple_mean_error = "simple")) %>%
  ggplot(aes(x = time, y = estimation_error)) + geom_line(aes(group = simulation_run, color = Method), alpha = 0.2) + 
  facet_wrap(~Method, ncol = 1) +
  custom_theme() + theme(strip.background = element_blank(), strip.text = element_blank()) + 
  scale_color_manual(values=cbPalette) +
  scale_fill_manual(values=cbPalette)

full_figure_top = ggarrange(panel_A, panel_B, panel_C, ncol=1, nrow = 3, align = "hv", labels=c("A", "B", "C"), common.legend = T)
full_figure_bottom = ggarrange(panel_D, labels=c("D"), common.legend = T)
ggarrange(full_figure_top, full_figure_bottom, ncol = 2, common.legend = T)

```

Often however, the the mean and standard deviation of the generation interval is typically not known exactly. For instance, for the COVID-19 the mean generation interval estimates, with 95% credible intervals, include 3.95 (3.01 - 4.91) and 5.20 (3.78 - 6.78); and estimates of the standard deviation of the generation interval include 1.51 (0.74 - 2.97) and 1.72 (0.91 - 3.93) [@ganyani_estimating_2020]. Therefore, we compared the accuracy of the two methods if the standard deviation of the generation interval is mis-specified as three days longer. Under these conditions, the simple ratio actually performs slightly better than than the Cori method (**Figure** \@ref(fig:estimation-inaccParams)).

@@

## Using the Simple Ratio to Estimate $R_t$ with real-world COVID-19 data {.unnumbered}

We next compared $R_t$ estimates from the Simple Ratio and Cori et al for using data from the current COVID-19 pandemic. We estimated $R_t$ in four nations using a generation interval with mean of `r COVID19_params$mean_gi` and standard deviation of `r COVID19_params$sd_gi`[@ganyani_estimating_2020]. **Figure** \@ref(fig:real-world-data) **A** shows the results for the month of October 2020. Visual inspection shows that the two methods of estimating $R_t$ have very good concordance. Across all nations over the course of the pandemic, the median difference in estimation between the two methods was `r estimation_diff_median` (95% range of `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`) (**Figure** \@ref(fig:real-world-data) **B**).

# Discussion {.unnumbered}

# References {.unnumbered}

::: {#refs}
:::

# Supplementary Material {.unnumbered}

\beginsupplement

```{=tex}
\newpage
\blandscape
```
```{r all-simulations, fig.width = 11, fig.height = 6, fig.cap = "All Simulations"}

rt_estimates %>%
  create_estimate_plot() + 
  facet_wrap(~simulation_run) + 
  theme(strip.background = element_blank(), strip.text = element_blank())
```

\elandscape

\newpage

```{r estimation-inaccParams,  fig.cap = "(A) estimation with gi mean mis-specified. (B) estimation with gi sd mis-specified."}
